Parallel Programming in Haskell Almost for Free

### 2013-10-06 src2/csort/Obsidian.tex
hacking

In the Obsidian project we experiment with combinators and language 
features for GPU kernel implementation. We seek to strike a balance 
between low-level control and useful high-level abstractions. Over time, this 
has led to a number of different versions of Obsidian. In reference \cite{JSLIC}, 
two versions of Obsidian differing by having a {\em monadic} or a limited {\em arrow} 
like interface are described. We have settled on using the monadic style.

@@

The Obsidian implementation of pull arrays was influenced by Pan's representation 
of images as functions from coordinates to colour values \cite{PAN}. A pull array 
is a function from index to value and an associated length.

Push arrays were added to Obsidian first in reference \cite{Obsidian-Expressive}. A push array
specifies where elements are to end up in a result array, rather than where 
they come from (as in a pull array). In essence, a push array is a
computation which writes an array to memory, but is parameterised on
the {\em write function} which writes a single element into memory.
The push array function can then 
use the write function zero, one, or several times.

Push arrays were added to Obsidian with the purpose of solving some performance 
issues that we had been struggling with for a long time. These problems were 
mainly concerned with concatenating or combining arrays in an efficient way.
Push arrays have also been adopted by others and are used in the implementation 
of filters in reference~\cite{FPCDSL} and in the new Nikola\footnote{github.com/mainland/nikola/blob/master/src/Data/Array/Nikola/Repr/Push.hs}.

@@

%% The push and pull arrays have different strengths. Pull arrays 
%% make it easy to implement functions such as {\tt map} and {\tt zipWith}. 
%% % and 
%% %also give map-fusion for free. A pull array is also easy to parallelise;  
%% %thread $i$ is in charge of computing element $i$.  
%% %The push arrays provide ways to compute arrays in a way so that thread $i$ 
%% %computes other elements than $i$. Actually, any thread be in charge of the 
%% %computation of any element of the result. 
%% Push arrays allow thread $i$ to compute array elements at indices other 
%% than $i$; this enables efficient array concatenation as well 
%% as the data scattering operations used later. For more information about 
%% push arrays, see reference \cite{PUSH}.

@@

%In the current version of Obsidian\footnote{github.com/SvenssonJoel/Obsidian} 
%we are exploring ways to increase the scope of what kinds of compute kernels 
%we can describe and obtain efficient code for. One aspect of this exploration 
%is the addition of a new kind of array representation that we call 
%{\em push} arrays \cite{PUSH}. Previous versions of Obsidian 
%(described in \cite{JSLIC}) represented arrays as functions 
%from indices to values,we call these {\em pull} arrays. Another addition 
%is global arrays. Earlier versions of Obsidian has been focused on 
%expressing the kind of computations that takes place locally, in a block.

@@

%We want to be able to sort large arrays and to do this we need an 
%implementation of prefix sum that is larger (on more elements) than what 
%fits in a block. 
We need to operate on arrays that are too long to fit in a single block.
%This large prefix sum is implemented as shown in
The resulting prefix sum is implemented as shown in reference 
\cite{ScanCUDA}, see figure \ref{fig:prefixsum}. There are three steps 
involved. First local prefix sums are computed; the rightmost elements elements are also stored in a separate, auxiliary, array. Following this, 
the prefix sum of the auxiliary array is  computed. The prefix sums used 
in phase A and B of the 
figure can both be generated from the same Obsidian program. Finally the 
local prefix sums and the result of the auxiliary prefix sum are combined, 
forming a large prefix sum over the entire array.

@@

The figure also shows what kernels we need to produce. First, something that 
performs a number of local prefix sums in parallel over the full length 
of an array is needed. The prefix sum algorithm that we implement here is 
based on divide and conquer and originates from Sklansky \cite{Sklansky}. 
% A kernel that performs the distribution (part C in the figure) is also needed.

@@

%To be able to implement prefix sum on the GPU we need to split the work 
%up into different phases. While computing a prefix sum in parallel , values 
%needs to be communicated between threads. This indicates that a kernel that 
%utilises local shared memory is a good idea. In CUDA the threads that can 
%communicate via shared memory all belong to the same block (group of threads). 
%These blocks can contain up to 1024 threads but often much fewer to 
%conserve resources and allow more blocks to run in parallel. In this example 
%a 256 threads (and elements) kernel is created. In the 
%benchmarks (section \ref{sec:Benchmarks}) prefix sum kernels of many 
%different sizes are used but all of these can be generated from the same 
%Obsidian description. In order to be able to implement a large prefix sum 
%the local kernel is also designed to output its maximal element to 
%a separate array. This separate array can be used to compute values to adjust 
%the local sums by. This follows the schema developed in \cite{LargeScan}.

>>>>

In the Obsidian project we experiment with combinators and language 
features for GPU kernel implementation. We seek to strike a balance 
between low-level control and useful high-level abstractions. Over time, this 
has led to a number of different versions of Obsidian. In reference \citecsort{JSLIC}, 
two versions of Obsidian differing by having a {\em monadic} or a limited {\em arrow} 
like interface are described. We have settled on using the monadic style.

@@

The Obsidian implementation of pull arrays was influenced by Pan's representation 
of images as functions from coordinates to colour values \citecsort{PAN}. A pull array 
is a function from index to value and an associated length.

Push arrays were added to Obsidian first in reference \citecsort{Obsidian-Expressive}. A push array
specifies where elements are to end up in a result array, rather than where 
they come from (as in a pull array). In essence, a push array is a
computation which writes an array to memory, but is parameterised on
the {\em write function} which writes a single element into memory.
The push array function can then 
use the write function zero, one, or several times.

Push arrays were added to Obsidian with the purpose of solving some performance 
issues that we had been struggling with for a long time. These problems were 
mainly concerned with concatenating or combining arrays in an efficient way.
Push arrays have also been adopted by others and are used in the implementation 
of filters in reference~\citecsort{FPCDSL} and in the new Nikola\footnote{github.com/mainland/nikola/blob/master/src/Data/Array/Nikola/Repr/Push.hs}.

@@

%% The push and pull arrays have different strengths. Pull arrays 
%% make it easy to implement functions such as {\tt map} and {\tt zipWith}. 
%% % and 
%% %also give map-fusion for free. A pull array is also easy to parallelise;  
%% %thread $i$ is in charge of computing element $i$.  
%% %The push arrays provide ways to compute arrays in a way so that thread $i$ 
%% %computes other elements than $i$. Actually, any thread be in charge of the 
%% %computation of any element of the result. 
%% Push arrays allow thread $i$ to compute array elements at indices other 
%% than $i$; this enables efficient array concatenation as well 
%% as the data scattering operations used later. For more information about 
%% push arrays, see reference \citecsort{PUSH}.

@@

%In the current version of Obsidian\footnote{github.com/SvenssonJoel/Obsidian} 
%we are exploring ways to increase the scope of what kinds of compute kernels 
%we can describe and obtain efficient code for. One aspect of this exploration 
%is the addition of a new kind of array representation that we call 
%{\em push} arrays \citecsort{PUSH}. Previous versions of Obsidian 
%(described in \citecsort{JSLIC}) represented arrays as functions 
%from indices to values,we call these {\em pull} arrays. Another addition 
%is global arrays. Earlier versions of Obsidian has been focused on 
%expressing the kind of computations that takes place locally, in a block.

@@

%We want to be able to sort large arrays and to do this we need an 
%implementation of prefix sum that is larger (on more elements) than what 
%fits in a block. 
We need to operate on arrays that are too long to fit in a single block.
%This large prefix sum is implemented as shown in
The resulting prefix sum is implemented as shown in reference 
\citecsort{ScanCUDA}, see figure \ref{fig:prefixsum}. There are three steps 
involved. First local prefix sums are computed; the rightmost elements elements are also stored in a separate, auxiliary, array. Following this, 
the prefix sum of the auxiliary array is  computed. The prefix sums used 
in phase A and B of the 
figure can both be generated from the same Obsidian program. Finally the 
local prefix sums and the result of the auxiliary prefix sum are combined, 
forming a large prefix sum over the entire array.

@@

The figure also shows what kernels we need to produce. First, something that 
performs a number of local prefix sums in parallel over the full length 
of an array is needed. The prefix sum algorithm that we implement here is 
based on divide and conquer and originates from Sklansky \citecsort{Sklansky}. 
% A kernel that performs the distribution (part C in the figure) is also needed.

@@

%To be able to implement prefix sum on the GPU we need to split the work 
%up into different phases. While computing a prefix sum in parallel , values 
%needs to be communicated between threads. This indicates that a kernel that 
%utilises local shared memory is a good idea. In CUDA the threads that can 
%communicate via shared memory all belong to the same block (group of threads). 
%These blocks can contain up to 1024 threads but often much fewer to 
%conserve resources and allow more blocks to run in parallel. In this example 
%a 256 threads (and elements) kernel is created. In the 
%benchmarks (section \ref{sec:Benchmarks}) prefix sum kernels of many 
%different sizes are used but all of these can be generated from the same 
%Obsidian description. In order to be able to implement a large prefix sum 
%the local kernel is also designed to output its maximal element to 
%a separate array. This separate array can be used to compute values to adjust 
%the local sums by. This follows the schema developed in \citecsort{LargeScan}.

### 2013-10-06 src2/csort/discussion.tex
apa
hacking

If the desired effect is to sort and remove duplicates 
(for an example of this see \cite{REMOVEDUPS}) then 
there is a clear benefit to using occurrence sort
over the {\tt sort} combined with {\tt unique} method.

In this paper, we are entirely focused on the implementation and
performance of kernels. Reference \cite{CSORT} goes into the details
of performing data transfers in parallel with ongoing computations. We
have not addressed this issue at all in our implementations. This is
also the reason why we haven't performed any benchmarks against the algorithm in
\cite{CSORT}, because it would compare very different things. As a
piece of future work, it would be interesting to add the capability of
streaming data from the main memory to the GPU to our algorithms.

In counting sort, an auxiliary array holding the count of how many times 
each element occurs in the input array, the histogram array, is created. 
The size of this array is 
based on the range of elements occurring in the input. This means that 
more memory is needed for larger ranges and may imply that counting sort 
is suitable only up to some point. A related algorithm, known as radix 
sort \cite{Knuth}, addresses this issue by using multiple counting sort 
like passes, one for each digit position and thus limits this auxiliary 
memory usage to an array of size related to the radix (number of unique 
digits) used in the representation of the values being sorted. During 
our exploration of the counting sort algorithm on a GPU, we were not much 
concerned by memory usage; all data and auxiliary 
arrays fit easily in the device memory. The charts (\ref{fig:chart1}, \ref{fig:chart3} and \ref{fig:chart2}) indicate that our implementation offers the most 
benefit in the ranges labelled as $T10$ to $T17$; these are in fact relatively 
small ranges (1024 to 131072 unique values). Our current hypothesis 
is that the decrease in benefit of counting sort from range $T18$ and onward
is due to more and more complicated memory access patterns in the 
reconstruction phase. However, more in-depth profiling is needed to 
understand where the decrease in benefit comes from.
%%  However, we postulate that 
%% the decrease in benefit following $T18$ and onward are due to more complicated
%% memory access patterns during reconstruction

@@

%% There is already an example of using Obsidian as the code 
%% generating back end of such an EDSL. This EDSL is called PEGGY and 
%% is described in \cite{cole2012beauty}.

>>>>

If the desired effect is to sort and remove duplicates 
(for an example of this see \citecsort{REMOVEDUPS}) then 
there is a clear benefit to using occurrence sort
over the {\tt sort} combined with {\tt unique} method.

In this paper, we are entirely focused on the implementation and
performance of kernels. Reference \citecsort{CSORT} goes into the details
of performing data transfers in parallel with ongoing computations. We
have not addressed this issue at all in our implementations. This is
also the reason why we haven't performed any benchmarks against the algorithm in
\citecsort{CSORT}, because it would compare very different things. As a
piece of future work, it would be interesting to add the capability of
streaming data from the main memory to the GPU to our algorithms.

In counting sort, an auxiliary array holding the count of how many times 
each element occurs in the input array, the histogram array, is created. 
The size of this array is 
based on the range of elements occurring in the input. This means that 
more memory is needed for larger ranges and may imply that counting sort 
is suitable only up to some point. A related algorithm, known as radix 
sort \citecsort{Knuth}, addresses this issue by using multiple counting sort 
like passes, one for each digit position and thus limits this auxiliary 
memory usage to an array of size related to the radix (number of unique 
digits) used in the representation of the values being sorted. During 
our exploration of the counting sort algorithm on a GPU, we were not much 
concerned by memory usage; all data and auxiliary 
arrays fit easily in the device memory. The charts (\ref{fig:chart1}, \ref{fig:chart3} and \ref{fig:chart2}) indicate that our implementation offers the most 
benefit in the ranges labelled as $T10$ to $T17$; these are in fact relatively 
small ranges (1024 to 131072 unique values). Our current hypothesis 
is that the decrease in benefit of counting sort from range $T18$ and onward
is due to more and more complicated memory access patterns in the 
reconstruction phase. However, more in-depth profiling is needed to 
understand where the decrease in benefit comes from.
%%  However, we postulate that 
%% the decrease in benefit following $T18$ and onward are due to more complicated
%% memory access patterns during reconstruction

@@

%% There is already an example of using Obsidian as the code 
%% generating back end of such an EDSL. This EDSL is called PEGGY and 
%% is described in \citecsort{cole2012beauty}.

### 2013-10-06 src2/csort/intro.tex
hacking

These algorithms are explored in Obsidian \cite{JSLIC,PUSH}, a domain
specific language targeting GPU programming. The goal of Obsidian is
to strike a balance between high-level constructs and low-level control. 
We want to provide the programmer with a set of tools for low-level 
experimentation with the details that influence performance when 
implementing GPU kernels. The counting sort
case study shows that Obsidian generates competitive kernels with
relatively little programmer effort. That Obsidian is an embedded language 
allows us to rapidly experiment with the addition of features and with varying 
programming idioms. The version used in this case study
adds global arrays and atomic instructions to Obsidian, see section 
\ref{sec:OBSHist}.

The contributions of this paper are:
\begin{itemize}
\item We provide measurements (section \ref{sec:CSORTBenchmarks}) showing
  that counting sort is a competitive algorithm for sorting keys on the GPU,
  outperforming the sorting implementation in the library
  Thrust\cite{THRUST} in many cases.
\item Occurrence sort is shown to be particularly
  suitable for implementing on the GPU (section \ref{sec:occur}) and
  performs well (section \ref{sec:CSORTBenchmarks}).
\item The Obsidian implementation of the two sorting algorithms is
  detailed along with the generated CUDA (sections
  \ref{sec:Obsidian} and \ref{sec:parallel}).
\end{itemize}

@@

Sorting has applications in the computer graphics field \cite{sintorn}. 
Example instances of sorting and duplicate element removal in computer graphics 
can be identified in references \cite{Olsson} and \cite{Karras}. Another example 
of where sorting and the removal of duplicates have applications is 
in databases \cite{REMOVEDUPS}.

%% Sorting on the GPU is interesting both for computer graphics
%% applications \cite{SINTORN} and for using the GPU to offload the
%% processor, for example in database applications \cite{REMOVEDUPS}.

The paper \cite{CSORT} implements counting sort in CUDA, and also optimisations 
that overlap computations with transferring data to and from the GPU.
Our work differs by being implemented in a high-level embedded
language and also by implementation of the occurrence sort variant of counting sort. We have not been concerned with
trying to overlap computations and data transfer, but have instead
focused solely on computations within the GPU.

Obsidian is a high-level embedded language for GPU
programming. Other such approaches include Accelerate
\cite{Acc} and Nikola \cite{NIKOLA}. Accelerate and Nikola both take an even
higher level approach compared to Obsidian and abstract more from GPU
details. Another example of providing a high-level interface to
programming the GPU is the C++ library Thrust \cite{THRUST}.
%\cite{FELDSPAR}

@@

%% This paper describes a case study. We implement a sorting algorithm known 
%% as counting sort. As an extra bonus we explore a variant of counting sort 
%% that can be given a nice purely functional description and lends itself well 
%% to parallelisation. The experimental results in section \ref{sec:Benchmarks} 
%% compares the performance of the counting sort algorithm we implement here to 
%% the NVIDIA Thrust \cite{THRUST} library. The Obsidian generated kernels perform
%% very well at relatively light work effort.

@@

%\cite{CSORT} 
%\cite{FELDSPAR}

>>>>

These algorithms are explored in Obsidian \citecsort{JSLIC,Obsidian-Expressive}, a domain
specific language targeting GPU programming. The goal of Obsidian is
to strike a balance between high-level constructs and low-level control. 
We want to provide the programmer with a set of tools for low-level 
experimentation with the details that influence performance when 
implementing GPU kernels. The counting sort
case study shows that Obsidian generates competitive kernels with
relatively little programmer effort. That Obsidian is an embedded language 
allows us to rapidly experiment with the addition of features and with varying 
programming idioms. The version used in this case study
adds global arrays and atomic instructions to Obsidian, see section 
\ref{sec:OBSHist}.

The contributions of this paper are:
\begin{itemize}
\item We provide measurements (section \ref{sec:CSORTBenchmarks}) showing
  that counting sort is a competitive algorithm for sorting keys on the GPU,
  outperforming the sorting implementation in the library
  Thrust\citecsort{THRUST} in many cases.
\item Occurrence sort is shown to be particularly
  suitable for implementing on the GPU (section \ref{sec:occur}) and
  performs well (section \ref{sec:CSORTBenchmarks}).
\item The Obsidian implementation of the two sorting algorithms is
  detailed along with the generated CUDA (sections
  \ref{sec:Obsidian} and \ref{sec:parallel}).
\end{itemize}

@@

Sorting has applications in the computer graphics field \citecsort{SINTORNSORT}. 
Example instances of sorting and duplicate element removal in computer graphics 
can be identified in references \citecsort{Olsson} and \citecsort{Karras}. Another example 
of where sorting and the removal of duplicates have applications is 
in databases \citecsort{REMOVEDUPS}.

%% Sorting on the GPU is interesting both for computer graphics
%% applications \citecsort{SINTORN} and for using the GPU to offload the
%% processor, for example in database applications \citecsort{REMOVEDUPS}.

The paper \citecsort{CSORT} implements counting sort in CUDA, and also optimisations 
that overlap computations with transferring data to and from the GPU.
Our work differs by being implemented in a high-level embedded
language and also by implementation of the occurrence sort variant of counting sort. We have not been concerned with
trying to overlap computations and data transfer, but have instead
focused solely on computations within the GPU.

Obsidian is a high-level embedded language for GPU
programming. Other such approaches include Accelerate
\citecsort{ACCELERATEDAMP11} and Nikola \citecsort{NIKOLA}. Accelerate and Nikola both take an even
higher level approach compared to Obsidian and abstract more from GPU
details. Another example of providing a high-level interface to
programming the GPU is the C++ library Thrust \citecsort{THRUST}.
%\citecsort{FELDSPAR}

@@

%% This paper describes a case study. We implement a sorting algorithm known 
%% as counting sort. As an extra bonus we explore a variant of counting sort 
%% that can be given a nice purely functional description and lends itself well 
%% to parallelisation. The experimental results in section \ref{sec:Benchmarks} 
%% compares the performance of the counting sort algorithm we implement here to 
%% the NVIDIA Thrust \citecsort{THRUST} library. The Obsidian generated kernels perform
%% very well at relatively light work effort.

@@

%\citecsort{CSORT} 
%\citecsort{FELDSPAR}

### 2013-10-06 src2/embarbb/implementation.tex
hacking

The {\tt d} parameter is used for a type level representation of the 
dimensionality of a vector, as in the Repa library~\cite{REPA}. The 
dimensionality is encoded using the following types together with 
the {\tt Int} type.

@@

The {\tt Expr} type also has constructors for the {\em call} and {\em map} 
functionality. To call a function means to apply it to input data. The 
{\tt map} operation specifies elementwise application of a function over 
vectors, like NESL's {\em apply-to-each}~\cite{NESL}, so it corresponds 
to Haskell's {\tt map} and {\tt zipWith}. Both {\tt Call} and {\tt Map} 
take a {\tt(R GenRecord)}. This is inspired by the delayed expressions 
that enable the implementation of {\tt vapply} in Nikola~\cite{NIKOLA}. 
The {\tt R} is the reification monad used to create DAGs (directed acyclic 
graphs) from embedded language functions. These DAGs are part of a 
{\tt GenRecord} that contains all information that the ArBB code generator 
needs to generate the ArBB function.

@@

The expression data type used in EmbArBB is untyped (not using a GADT). 
A typed interface to the language is supplied using the same phantom types 
method as used in {\em pan}~\cite{ELLIJFP} and many other Haskell embedded 
languages since then. The choice to use an untyped {\tt Expr} datatype was 
based on the wish to keep the backend (ArBB code generation) as simple as 
possible.

@@

When a function {\tt f} of type 
\begin{verbatim}
Exp tin1 -> ... -> Exp tinN -> Exp tout
\end{verbatim}
is captured, it is 
first applied to expressions that represent variable names. For each of the 
inputs, {\tt (tin1,..,tinN}), a variable is created. The result is an expression 
(or expressions) representing the function {\tt f}. On this expression, sharing 
detection is performed, and a directed acyclic graph (DAG) is created. 
The method of sharing detection used is based on the StableNames method~\cite{Gill}.

>>>>

The {\tt d} parameter is used for a type level representation of the 
dimensionality of a vector, as in the Repa library~\citeemb{REPA}. The 
dimensionality is encoded using the following types together with 
the {\tt Int} type.

@@

The {\tt Expr} type also has constructors for the {\em call} and {\em map} 
functionality. To call a function means to apply it to input data. The 
{\tt map} operation specifies elementwise application of a function over 
vectors, like NESL's {\em apply-to-each}~\citeemb{NESL}, so it corresponds 
to Haskell's {\tt map} and {\tt zipWith}. Both {\tt Call} and {\tt Map} 
take a {\tt(R GenRecord)}. This is inspired by the delayed expressions 
that enable the implementation of {\tt vapply} in Nikola~\citeemb{NIKOLA}. 
The {\tt R} is the reification monad used to create DAGs (directed acyclic 
graphs) from embedded language functions. These DAGs are part of a 
{\tt GenRecord} that contains all information that the ArBB code generator 
needs to generate the ArBB function.

@@

The expression data type used in EmbArBB is untyped (not using a GADT). 
A typed interface to the language is supplied using the same phantom types 
method as used in {\em pan}~\citeemb{COMPILEEDSL} and many other Haskell embedded 
languages since then. The choice to use an untyped {\tt Expr} datatype was 
based on the wish to keep the backend (ArBB code generation) as simple as 
possible.

@@

When a function {\tt f} of type 
\begin{verbatim}
Exp tin1 -> ... -> Exp tinN -> Exp tout
\end{verbatim}
is captured, it is 
first applied to expressions that represent variable names. For each of the 
inputs, {\tt (tin1,..,tinN}), a variable is created. The result is an expression 
(or expressions) representing the function {\tt f}. On this expression, sharing 
detection is performed, and a directed acyclic graph (DAG) is created. 
The method of sharing detection used is based on the StableNames method~\citeemb{Gill}.

### 2013-10-06 src2/embarbb/introduction.tex
hacking

Intel's Array Building Blocks (ArBB) is one approach to the problem of how to increase productivity for programmers who need to exploit hardware parallelism, but for whom very low level parallel programming (as typically found in High Performance Computing) is too difficult or time consuming~\cite{ARBB2011}. ArBB is a retargettable dynamic compilation framework that is provided as an embedded language in C++. It aims to give the programmer access to data and thread parallelism, while protecting him from common pitfalls such as race conditions and deadlock.

Many of the ideas in ArBB feel familiar to functional programmers.
For example, when referring to collection data types, reference~\cite{ARBB2011} states that ``Expressions over collections always act as if a completely new collection were created and assignment between collections always behaves as if the entire input collection were copied.''
This rather functional view of data is accompanied by extensive optimisation that (as the paper states) removes almost all copying.
A key to the approach is the use of standard patterns of computation such as scan and reduce -- familiar as higher order functions in a functional setting.
The system tries to fuse away intermediate data-structures
when these patterns are composed.
The patterns are deterministic, so that problems with race conditions are avoided by design. Programs are stated to be short and close to the mathematical specification, and that, again, is a familiar mantra. The system allows C++ users to construct code objects called closures.
The first time the {\em call} operation is applied to
a particular function, ArBB captures the sequence of operations
over ArBB types and creates an intermediate representation. 
This representation is then compiled to
vector machine language, which is
cached for later use.  The
machine code is also invoked (in parallel on multiple
cores, if appropriate). The next time the function
is called, ArBB reuses the cached machine
code rather than regenerating it.
Although dynamically typed closures are available, ArBB is generally
statically typed.

@@

%ArBB is programming language that aspires to
%make it easier to program modern multi-core CPUs. ArBB gives the 
%programmer a set of operations that when used results in automatic 
%vectorisation and threading. 
%Amongst these operations are scans (prefix sums)  
%and reductions over vectors, elementwise operations and more. 
%The compilation 
%process also tries to fuse away intermediate datastructures~\cite{ARBB2011}.

ArBB is an embedded language, implemented
as a library in the host language C++. The embedding makes 
heavy use of the C++ template machinery. There is also a lower level 
interface to ArBB, the ArBB Virtual Machine C API~\cite{arbbvm}. The low level C API is not
intended for use directly by applications programmers, but rather as a base 
for implementations of alternative ArBB frontends (embeddings)~\cite{ARBB2011}.
The ArBB VM implements the compilation and parallelisation.
Reference~\cite{joelryan} presents Haskell bindings to the ArBB C API as well as initial 
steps towards using ArBB as a backend to the embedded language Accelerate~\cite{ACCELERATEDAMP11}.

>>>>

Intel's Array Building Blocks (ArBB) is one approach to the problem of how to increase productivity for programmers who need to exploit hardware parallelism, but for whom very low level parallel programming (as typically found in High Performance Computing) is too difficult or time consuming~\citeemb{ARBB2011}. ArBB is a retargettable dynamic compilation framework that is provided as an embedded language in C++. It aims to give the programmer access to data and thread parallelism, while protecting him from common pitfalls such as race conditions and deadlock.

Many of the ideas in ArBB feel familiar to functional programmers.
For example, when referring to collection data types, reference~\citeemb{ARBB2011} states that ``Expressions over collections always act as if a completely new collection were created and assignment between collections always behaves as if the entire input collection were copied.''
This rather functional view of data is accompanied by extensive optimisation that (as the paper states) removes almost all copying.
A key to the approach is the use of standard patterns of computation such as scan and reduce -- familiar as higher order functions in a functional setting.
The system tries to fuse away intermediate data-structures
when these patterns are composed.
The patterns are deterministic, so that problems with race conditions are avoided by design. Programs are stated to be short and close to the mathematical specification, and that, again, is a familiar mantra. The system allows C++ users to construct code objects called closures.
The first time the {\em call} operation is applied to
a particular function, ArBB captures the sequence of operations
over ArBB types and creates an intermediate representation. 
This representation is then compiled to
vector machine language, which is
cached for later use.  The
machine code is also invoked (in parallel on multiple
cores, if appropriate). The next time the function
is called, ArBB reuses the cached machine
code rather than regenerating it.
Although dynamically typed closures are available, ArBB is generally
statically typed.

@@

%ArBB is programming language that aspires to
%make it easier to program modern multi-core CPUs. ArBB gives the 
%programmer a set of operations that when used results in automatic 
%vectorisation and threading. 
%Amongst these operations are scans (prefix sums)  
%and reductions over vectors, elementwise operations and more. 
%The compilation 
%process also tries to fuse away intermediate datastructures~\citeemb{ARBB2011}.

ArBB is an embedded language, implemented
as a library in the host language C++. The embedding makes 
heavy use of the C++ template machinery. There is also a lower level 
interface to ArBB, the ArBB Virtual Machine C API~\citeemb{arbbvm}. The low level C API is not
intended for use directly by applications programmers, but rather as a base 
for implementations of alternative ArBB frontends (embeddings)~\citeemb{ARBB2011}.
The ArBB VM implements the compilation and parallelisation.
Reference~\citeemb{joelryan} presents Haskell bindings to the ArBB C API as well as initial 
steps towards using ArBB as a backend to the embedded language Accelerate~\citeemb{ACCELERATEDAMP11}.

### 2013-10-06 src2/embarbb/related.tex
apa
hacking

\subsection{Related Work} 
\subsubsection{Data.Array.Accelerate} 
Accelerate~\cite{ACCELERATEDAMP11} is an embedded language for general purpose, 
flat data parallel computations on GPUs.

@@

DPH relies on a technique called flattening to transform its nested data parallelism 
into flat data parallelism. 
A recent article about DPH is reference~\cite{DPH}.

\subsubsection{Feldspar} 
Feldspar is a language for Digital Signal Processing (DSP) programming
developed at Chalmers, ELTE University and Ericsson 
(the telecom company)~\cite{FELDSPAR2010}. The functional {\tt while} loop 
of EmbArBB is inspired by the same concept in Feldspar.

@@

%The most recent release of Feldspar uses the {\em Syntactic} library
%for manipulating abstract syntax~\cite{SyntacticICFP12,FELDSYNTACTIC}. 
%We will consider switching to using the Syntactic library in the continued
%implementation of EmbArbb.
%Initial discussions with our colleague Emil Axelsson, the implementor of Syntactic, indicate that
%we should encounter no major difficulties.

\subsubsection{Nikola} 
Nikola is an embedded language for GPU programming\cite{NIKOLA}, also
in Haskell.  
During the early phases of implementation of EmbArBB, the source code of 
Nikola was often studied for inspiration.

The embedding used in Nikola makes use of an untyped expression 
data structure wrapped up in phantom types, in the style of Pan~\cite{ELLIJFP}.
The EmbArBB embedding works in the same way. 
One of the listed strengths of Nikola is the ability to generate GPU functions 
from Haskell functions, which enables function reuse and the ability to 
amortise the cost of code generation over several launches in an easy 
way. The ability the generate target language function from Haskell functions 
is a present in EmbArBB as well.

@@

\subsubsection{Repa} 
Repa is a library for regular, shape polymorphic parallel arrays~\cite{REPA}. Repa 
uses a concept of {\tt delayed} arrays to obtain fusion of operations, such as the 
typical map fusion:
\begin{verbatim}
map f . map g  = map (f . g)
\end{verbatim} 
A delayed array 
has a representation that is quite direct to parallelise; it is represented 
as a function from an index-space to an element and the extents of that same 
index-space. Concretely:

@@

%\subsection{Obsidian}
%Really ? \cite{Obsidian}

@@

Firstly, we are interested in exploring programming idioms for
data parallel programming in a functional setting. Embedding ArBB gives
a quick route to a platform for such exploration, without too much implementation effort. This is what the ``almost for free'' in the title is intended to refer to.
We wish to explore ways to make use of the fact that we have an array programming library embedded in a sophisticated, strongly typed host language. In our earlier work on the embedded hardware description language Lava, we investigated various approaches to exploiting the host language during netlist synthesis~\cite{LavaMultipliers}, and we have also experimented with the use of search and dynamic programming in generating parallel prefix (or scan) networks~\cite{SheeranJFP}. We intend to apply similar methods to the development of data parallel programs once the EmbArBB implementation is more complete and stable.

%It would be interesting to explore ways of using the host language in more sophisticated ways, rather as we have done for hardware generation in Lava~\cite{LavaMultipliers}.

A second motivation is the desire to teach NESL-style data parallel programming
in a Masters course on parallel functional programming that has recently been
introduced at Chalmers~\cite{TFPIEPFP}. The first instance of the course (in Spring 2012)
covered Blelloch's NESL language, with its
associated cost model~\cite{NESL}, but did not provide any satisfactory way for the students
to experience real, nested data parallel programming. This was due not only to a lack of suitable hardware, but also to deficiencies in the available tools.
We used the Repa library to get flat data parallelism, but then suffered from
the lack of a built-in scan primitive (as many of Blelloch's NESL examples use scan). Perhaps as a result
of needing scan, we did not get good performance.
We have not done enough examples or experiments yet, but it seems to us that EmbArBB
could be used to give students an experience of real parallel programming in a NESL-like
functional language. It remains to be seen whether the limited degree of nesting allowed in ArBB can be offset, at least to some extent, by clever use of the host programming language during generation of the desired abstract syntax tree.

>>>>

\subsection{Related Work} 
\subsubsection{Data.Array.Accelerate} 
Accelerate~\citeemb{ACCELERATEDAMP11} is an embedded language for general purpose, 
flat data parallel computations on GPUs.

@@

DPH relies on a technique called flattening to transform its nested data parallelism 
into flat data parallelism. 
A recent article about DPH is reference~\citeemb{DPH}.

\subsubsection{Feldspar} 
Feldspar is a language for Digital Signal Processing (DSP) programming
developed at Chalmers, ELTE University and Ericsson 
(the telecom company)~\citeemb{FELDSPAR2010}. The functional {\tt while} loop 
of EmbArBB is inspired by the same concept in Feldspar.

@@

%The most recent release of Feldspar uses the {\em Syntactic} library
%for manipulating abstract syntax~\citeemb{SyntacticICFP12,FELDSYNTACTIC}. 
%We will consider switching to using the Syntactic library in the continued
%implementation of EmbArbb.
%Initial discussions with our colleague Emil Axelsson, the implementor of Syntactic, indicate that
%we should encounter no major difficulties.

\subsubsection{Nikola} 
Nikola is an embedded language for GPU programming\citeemb{NIKOLA}, also
in Haskell.  
During the early phases of implementation of EmbArBB, the source code of 
Nikola was often studied for inspiration.

The embedding used in Nikola makes use of an untyped expression 
data structure wrapped up in phantom types, in the style of Pan~\citeemb{COMPILEEDSL}.
The EmbArBB embedding works in the same way. 
One of the listed strengths of Nikola is the ability to generate GPU functions 
from Haskell functions, which enables function reuse and the ability to 
amortise the cost of code generation over several launches in an easy 
way. The ability the generate target language function from Haskell functions 
is a present in EmbArBB as well.

@@

\subsubsection{Repa} 
Repa is a library for regular, shape polymorphic parallel arrays~\citeemb{REPA}. Repa 
uses a concept of {\tt delayed} arrays to obtain fusion of operations, such as the 
typical map fusion:
\begin{verbatim}
map f . map g  = map (f . g)
\end{verbatim} 
A delayed array 
has a representation that is quite direct to parallelise; it is represented 
as a function from an index-space to an element and the extents of that same 
index-space. Concretely:

@@

%\subsection{Obsidian}
%Really ? \citeemb{Obsidian}

@@

Firstly, we are interested in exploring programming idioms for
data parallel programming in a functional setting. Embedding ArBB gives
a quick route to a platform for such exploration, without too much implementation effort. This is what the ``almost for free'' in the title is intended to refer to.
We wish to explore ways to make use of the fact that we have an array programming library embedded in a sophisticated, strongly typed host language. In our earlier work on the embedded hardware description language Lava, we investigated various approaches to exploiting the host language during netlist synthesis~\citeemb{LavaMultipliers}, and we have also experimented with the use of search and dynamic programming in generating parallel prefix (or scan) networks~\citeemb{SheeranJFP}. We intend to apply similar methods to the development of data parallel programs once the EmbArBB implementation is more complete and stable.

%It would be interesting to explore ways of using the host language in more sophisticated ways, rather as we have done for hardware generation in Lava~\citeemb{LavaMultipliers}.

A second motivation is the desire to teach NESL-style data parallel programming
in a Masters course on parallel functional programming that has recently been
introduced at Chalmers~\citeemb{TFPIEPFP}. The first instance of the course (in Spring 2012)
covered Blelloch's NESL language, with its
associated cost model~\citeemb{NESL}, but did not provide any satisfactory way for the students
to experience real, nested data parallel programming. This was due not only to a lack of suitable hardware, but also to deficiencies in the available tools.
We used the Repa library to get flat data parallelism, but then suffered from
the lack of a built-in scan primitive (as many of Blelloch's NESL examples use scan). Perhaps as a result
of needing scan, we did not get good performance.
We have not done enough examples or experiments yet, but it seems to us that EmbArBB
could be used to give students an experience of real parallel programming in a NESL-like
functional language. It remains to be seen whether the limited degree of nesting allowed in ArBB can be offset, at least to some extent, by clever use of the host programming language during generation of the desired abstract syntax tree.
